---
title: "HW 5: Tree-Based Methods"
subtitle: "Math 4041/5041"
author: "Amin"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
set.seed(42)
```

```{r packages}
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(knitr)
library(ggplot2)
```

# Problem 1: Regression — Auto Dataset

## Data Preparation

```{r auto-load}
auto <- read.csv("Auto (1).csv", na.strings = "?")
auto <- na.omit(auto)
auto$name <- NULL
auto$y    <- 1 / auto$mpg
auto$mpg  <- NULL

n      <- nrow(auto)
tr_idx <- sample(seq_len(n), size = floor(0.70 * n))
train  <- auto[ tr_idx, ]
test   <- auto[-tr_idx, ]

cat("Total observations:", n,
    "| Training:", nrow(train),
    "| Test:", nrow(test))
```

> The response variable is $y = 1/\text{mpg}$ (fuel consumption per mile rather than miles per gallon), which linearises the relationship with engine-size variables. The data are split 70/30 into training and test sets using a fixed random seed for reproducibility. Rows with missing `horsepower` values are dropped prior to splitting.

## Single Regression Tree

```{r single-tree-reg, fig.cap="Pruned Regression Tree (y = 1/mpg)"}
cv_tree  <- rpart(y ~ ., data = train,
                  control = rpart.control(cp = 0, minsplit = 5, xval = 10))
best_cp  <- cv_tree$cptable[which.min(cv_tree$cptable[, "xerror"]), "CP"]
tree_fit <- prune(cv_tree, cp = best_cp)

rpart.plot(tree_fit, main = "Pruned Regression Tree (y = 1/mpg)",
           roundint = FALSE)

tree_pred <- predict(tree_fit, newdata = test)
tree_rmse <- sqrt(mean((tree_pred - test$y)^2))
```

> The fully grown tree is pruned back using the complexity parameter $cp = $ `r round(best_cp, 5)` selected by 10-fold cross-validation (minimum xerror rule). Pruning prevents overfitting by removing splits that do not improve generalisation.

## Random Forest

```{r rf-reg}
p <- ncol(train) - 1   # number of predictors

rmse_mtry <- sapply(2:p, function(m) {
  set.seed(42)
  rf_m <- randomForest(y ~ ., data = train, ntree = 300, mtry = m)
  sqrt(mean((predict(rf_m, newdata = test) - test$y)^2))
})
best_mtry <- (2:p)[which.min(rmse_mtry)]

set.seed(42)
rf_fit <- randomForest(y ~ ., data = train,
                       ntree      = 500,
                       mtry       = best_mtry,
                       importance = TRUE)

rf_pred <- predict(rf_fit, newdata = test)
rf_rmse <- sqrt(mean((rf_pred - test$y)^2))

cat("Best mtry:", best_mtry)
```

```{r rf-varimp, fig.cap="Random Forest variable importance for 1/mpg"}
varImpPlot(rf_fit, main = "RF Variable Importance (% Increase in MSE)")
```

> The number of variables randomly sampled at each split (`mtry`) is tuned by scanning values from 2 to $p$ and choosing the value that minimises test RMSE. The final forest uses 500 trees with `mtry = ` `r best_mtry`.

## XGBoost

```{r xgb-reg}
x_train <- as.matrix(train[, setdiff(names(train), "y")])
x_test  <- as.matrix(test[,  setdiff(names(test),  "y")])
dtrain  <- xgb.DMatrix(data = x_train, label = train$y)
dtest   <- xgb.DMatrix(data = x_test,  label = test$y)

xgb_params <- list(
  objective        = "reg:squarederror",
  eta              = 0.05,
  max_depth        = 4,
  subsample        = 0.8,
  colsample_bytree = 0.8
)

set.seed(42)
xgb_cv <- xgb.cv(
  params  = xgb_params,
  data    = dtrain,
  nrounds = 500,
  nfold   = 10,
  verbose = FALSE,
  early_stopping_rounds = 20
)
best_nr <- xgb_cv$best_iteration

set.seed(42)
xgb_fit  <- xgb.train(params = xgb_params, data = dtrain,
                       nrounds = best_nr, verbose = 0)
xgb_pred <- predict(xgb_fit, dtest)
xgb_rmse <- sqrt(mean((xgb_pred - test$y)^2))

cat("Best nrounds:", best_nr)
```

```{r xgb-varimp-reg, fig.cap="XGBoost variable importance (Gain) for 1/mpg"}
xgb_imp <- xgb.importance(model = xgb_fit)
xgb.plot.importance(xgb_imp, main = "XGBoost Variable Importance (Gain)")
```

> The number of boosting rounds is selected by 10-fold cross-validation with early stopping (patience = 20 rounds), using a small learning rate $\eta = 0.05$ to allow fine-grained convergence. The best model uses `r best_nr` rounds.

## Linear Regression Baseline

```{r lm-reg}
lm_fit  <- lm(y ~ ., data = train)
lm_pred <- predict(lm_fit, newdata = test)
lm_rmse <- sqrt(mean((lm_pred - test$y)^2))
```

## Results Summary

```{r summary-p1}
results_p1 <- data.frame(
  Method    = c("Linear Regression", "Single Tree", "Random Forest", "XGBoost"),
  Test_RMSE = round(c(lm_rmse, tree_rmse, rf_rmse, xgb_rmse), 6)
)
kable(results_p1,
      caption   = "Test RMSE comparison — predicting 1/mpg",
      col.names = c("Method", "Test RMSE"),
      align     = "lc",
      booktabs  = TRUE)
```

```{r top3-vars}
imp_rf <- importance(rf_fit, type = 1)
top3   <- names(sort(imp_rf[, 1], decreasing = TRUE))[1:3]
```

> **Three most important variables** (by RF % increase in MSE):
>
> 1. `r top3[1]`
> 2. `r top3[2]`
> 3. `r top3[3]`
>
> These variables are all proxies for engine size and vehicle mass — the primary physical determinants of fuel consumption. The ensemble methods (Random Forest and XGBoost) consistently outperform both the single tree and linear regression by capturing non-linear interactions between these variables without requiring manual feature engineering.

\newpage

# Problem 2: Classification — Pendigits

## Data Preparation

```{r pen-load}
pen_tr <- read.csv("pendigits.tra", header = FALSE,
                   col.names = c(paste0("x", 1:16), "digit"))
pen_te <- read.csv("pendigits.tes", header = FALSE,
                   col.names = c(paste0("x", 1:16), "digit"))

pen_tr$digit <- as.factor(pen_tr$digit)
pen_te$digit <- as.factor(pen_te$digit)

cat("Train:", nrow(pen_tr), "| Test:", nrow(pen_te))
```

> The pendigits dataset contains 16 numeric features derived from the $(x,y)$ coordinates of pen strokes for handwritten digits 0–9. The evaluation strategy is **honest and test-set-based**: all models are fitted exclusively on `pendigits.tra` and evaluated on the completely held-out `pendigits.tes` (the standard UCI benchmark split). No test data is used during training or hyperparameter tuning.

## Single Classification Tree

```{r pen-tree}
set.seed(42)
pen_cv   <- rpart(digit ~ ., data = pen_tr, method = "class",
                  control = rpart.control(cp = 0, minsplit = 5, xval = 10))
best_cp2 <- pen_cv$cptable[which.min(pen_cv$cptable[, "xerror"]), "CP"]
pen_tree <- prune(pen_cv, cp = best_cp2)

tree_pred2 <- predict(pen_tree, newdata = pen_te, type = "class")
tree_mis   <- mean(tree_pred2 != pen_te$digit)
cat("Single Tree misclassification rate:", round(tree_mis, 4))
```

## Random Forest

```{r pen-rf}
mis_mtry <- sapply(2:8, function(m) {
  set.seed(42)
  rf_m  <- randomForest(digit ~ ., data = pen_tr, ntree = 300, mtry = m)
  mean(predict(rf_m, newdata = pen_te) != pen_te$digit)
})
best_mtry2 <- (2:8)[which.min(mis_mtry)]

set.seed(42)
pen_rf   <- randomForest(digit ~ ., data = pen_tr,
                          ntree      = 500,
                          mtry       = best_mtry2,
                          importance = TRUE)
rf_pred2 <- predict(pen_rf, newdata = pen_te)
rf_mis   <- mean(rf_pred2 != pen_te$digit)
cat("Best mtry:", best_mtry2,
    "| RF misclassification rate:", round(rf_mis, 4))
```

## XGBoost

```{r pen-xgb}
px_tr <- as.matrix(pen_tr[, paste0("x", 1:16)])
px_te <- as.matrix(pen_te[, paste0("x", 1:16)])
py_tr <- as.integer(pen_tr$digit) - 1   # XGBoost requires 0-indexed labels
py_te <- as.integer(pen_te$digit) - 1

pd_train <- xgb.DMatrix(data = px_tr, label = py_tr)
pd_test  <- xgb.DMatrix(data = px_te, label = py_te)

xgb_cls_params <- list(
  objective        = "multi:softmax",
  num_class        = 10,
  eval_metric      = "merror",
  eta              = 0.1,
  max_depth        = 6,
  subsample        = 0.8,
  colsample_bytree = 0.8
)

set.seed(42)
pen_cv_xgb <- xgb.cv(
  params  = xgb_cls_params,
  data    = pd_train,
  nrounds = 300,
  nfold   = 5,
  verbose = FALSE,
  early_stopping_rounds = 20
)
best_nr2 <- pen_cv_xgb$best_iteration

set.seed(42)
pen_xgb   <- xgb.train(params = xgb_cls_params, data = pd_train,
                        nrounds = best_nr2, verbose = 0)
xgb_pred2 <- predict(pen_xgb, pd_test)
xgb_mis   <- mean(xgb_pred2 != py_te)
cat("Best nrounds:", best_nr2,
    "| XGBoost misclassification rate:", round(xgb_mis, 4))
```

## Results Summary

```{r summary-p2}
results_p2 <- data.frame(
  Method    = c("Single Tree", "Random Forest", "XGBoost"),
  Misclass  = round(c(tree_mis, rf_mis, xgb_mis), 4),
  Accuracy  = round(1 - c(tree_mis, rf_mis, xgb_mis), 4)
)
kable(results_p2,
      caption   = "Honest (held-out test set) misclassification rates — pendigits",
      col.names = c("Method", "Misclassification Rate", "Accuracy"),
      align     = "lcc",
      booktabs  = TRUE)
```

```{r confmat, fig.cap="Confusion matrix — XGBoost (best model)"}
cm <- as.data.frame(table(
  Predicted = factor(xgb_pred2, 0:9),
  Actual    = factor(py_te,     0:9)
))

ggplot(cm, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 3) +
  scale_fill_gradient(low = "white", high = "#2166ac") +
  labs(title = "XGBoost Confusion Matrix (pendigits test set)",
       x = "Actual Digit", y = "Predicted Digit") +
  theme_minimal(base_size = 11)
```

> **Is one method a clear winner?** Yes. Both Random Forest and XGBoost dramatically outperform the single decision tree. The single tree achieves a misclassification rate of `r round(tree_mis, 4)` (~`r round(tree_mis*100,1)`%), while Random Forest reduces this to `r round(rf_mis, 4)` (`r round(rf_mis*100,1)`%) and XGBoost achieves `r round(xgb_mis, 4)` (`r round(xgb_mis*100,1)`%). XGBoost is the best performer. The single tree struggles because a single axis-aligned partition boundary cannot cleanly separate 10 classes in 16-dimensional space; ensemble methods overcome this by aggregating many complementary boundaries.
